---
title: "Assignment 1"
author: "Enya Liu"
date: "2025-03-24"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(tidyverse)
```

```{r}
# read datasets
registry <- read.csv("C:/Users/umi/Downloads/registry_deID.csv")
```

# Task 1 Creating a summarizing registry dataset
## Check for participants who did not prescirbe any modulator
```{r}
# process the registry dataset first
# replace NA with 0
registry[is.na(registry)] <- 0
no_registry <- registry %>%
  group_by(newID) %>%
  summarise(Vx770 = mean(Vx770), Vx809 = mean(Vx809comb), vx661 = mean(vx661comb), vx445 = mean(vx445comb))

no_registry$is_all_zero <- rowSums(no_registry[, c("Vx770", "Vx809", "vx661", "vx445")] == 0) == 4
```

```{r}
# check for columns and convert them into a single group column with encounter or prescription groups
edited_registry <- registry %>%
  mutate(
    group = ifelse(
      rowSums(select(., Vx770, Vx809comb, vx661comb, vx445comb)) == 0,
      "encounter",
      apply(select(., Vx770, Vx809comb, vx661comb, vx445comb), 1, 
            function(row){paste(names(which(row == 1)), collapse = ", ")})
    )
  )

# if one date has multiple prescriptions, separate them into separate rows
edited_registry <- edited_registry %>% separate_rows(group, sep = ", ")

# keep encounter date as a separate dataframe and compare it to the edited registry data later
encounter_registry <- edited_registry[edited_registry$group=="encounter",]
encounter_registry <- encounter_registry %>%
  group_by(newID) %>%
# [MCM Comment] The issue begins here. You are defining the first/final encounter dates using
# only dates during which participants were not prescribed a modulator. What if the participant
# was on a modulator at the time of the first/last registry entry?
  summarise(first_enc_date = min(stdyrs),
    final_enc_date = max(stdyrs))

edited_registry <- edited_registry[!edited_registry$group=="encounter",]

# group by ID and prescription group and find the earliest and latest prescribed date
edited_registry <- edited_registry %>%
  group_by(newID, group) %>%
  summarise(start_date = min(stdyrs),
    stop_date = max(stdyrs))

# add encounter date back by full join
final_registry_data <- edited_registry %>%
  full_join(encounter_registry, by = c("newID" = "newID"))

# if no prescription
final_registry_data <- final_registry_data %>%
  mutate(
    start_date = if_else(is.na(start_date), 0, start_date),
    stop_date = if_else(is.na(stop_date), 0, stop_date)
  )


# check if stop date > final enc date
final_registry_data$final_enc_date <- ifelse(final_registry_data$stop_date > final_registry_data$final_enc_date, 
                                             final_registry_data$stop_date, final_registry_data$final_enc_date)

# Use the latest encounter date for each participant
final_registry_data <- final_registry_data %>%
  group_by(newID) %>%
  mutate(final_enc_date = max(final_enc_date)) %>%
  ungroup()

# replace vx groups with prescribe names
final_registry_data <- final_registry_data %>%
  mutate(group = recode(group,
                         "Vx770" = "Ivacaftor",
                         "Vx809comb" = "Lumacaftor/Ivacaftor",
                         "vx661comb" = "Tezacaftor/Ivacaftor",
                        "vx445comb" = "Elexacafot/Tezacaftor/Ivacaftor"))
```

# Task 2 Create an indicator variable status
```{r}
final_registry_data$status <- ifelse(final_registry_data$stop_date == final_registry_data$final_enc_date, 0, 1)
# show result
final_registry_data
```

# Task 3 Save the dataset as CHEC_tx_dates.csv
```{r}
write.csv(final_registry_data, "CHEC_tx_dates.csv")
```
The processed registry data was saved as CHEC_tx_dates.csv to the working directory. 

# Task 4 Histogram Visualisation 
```{r}
histogram_data <- final_registry_data %>% group_by(group, newID) %>%
  summarise(value = abs(start_date-final_enc_date))

# Compute summary statistics by group
summary_stats <- histogram_data %>%
  group_by(group) %>%
  summarize(
    mean = mean(value, na.rm = T),
    min = min(value, na.rm = T),
    q1 = quantile(value, 0.25, na.rm = T),
    median = median(value, na.rm = T),
    q3 = quantile(value, 0.75, na.rm = T),
    max = max(value, na.rm = T)
  ) %>%
  tidyr::pivot_longer(cols = -group, names_to = "stat", values_to = "xintercept")

ggplot(histogram_data, aes(x = value, fill = group)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 30) +
  facet_wrap(~group) +
  labs(title = "Histogram Stratified by Group", x = "Value", y = "Count") +
  geom_vline(data = summary_stats, aes(xintercept = xintercept, color = stat))
```
# Task 5 Number of Participants Prescribed CFTR Modulator
```{r}
cftr_participants <- final_registry_data %>%
  group_by(group) %>%
  summarise(n_distinct(newID))
cftr_participants
```
# Task 6 Number of CFTR Modulators Prescribed
```{r}
final_registry_data <- final_registry_data[!is.na(final_registry_data$group),]
num_CFTR <- final_registry_data %>%
  group_by(newID) %>%
  summarise(n_distinct(group))
count(num_CFTR[num_CFTR$`n_distinct(group)`== 1,])
count(num_CFTR[num_CFTR$`n_distinct(group)`== 2,])
count(num_CFTR[num_CFTR$`n_distinct(group)`== 3,])
count(num_CFTR[num_CFTR$`n_distinct(group)`== 4,])
```


